Quantifying Limits to Detection of Early Warning for Critical Transitions

### 2012-04-21 ReplyToReviewers.md

Replies in _italics_

Referee: 1
==========

### Comments to the Author
I think that this is one of the most important papers yet written in what is emerging as an area of great interest.  My only concern is that it is written in a highly technical way that may not be accessible to the broad audience that should read it.   The idea of using a model-based approach and approaching the problem as one of model choice is very clever, but I had to read the section three times before I knew what that meant.   I would move some of that material up front to make the main points a little less drily than they are currently made. I think also some technical terms should be embellished with discussion and explanation up front.

_We thank the reviewer for this assessment, and we have done our best to address this important issue_

Tell the reader, for example, what a saddle-node bifurcation is.

_A definition now follows the first mention_

Explain how the coefficient of the noise term in the stochastic differential equation is chosen.

_Certainly this needed more explanation for any audience.  The justification now follows where this appears_

This part is written for experts, whereas for Interfaces it needs to be more pedagogical.

Why does the bifurcation parameter have a subscript t?

_to indicate it is changing in time, this has been stated when it first appears_

### Small points:  
- Line 25:  lead should be led
- Line 298:  Data are
- Line 309:  its

_Indeed, thanks for these corrections._

I like the examples.

Referee: 2
Comments to the Author
I am pleased to recommend this manuscript for publication in Proceedings of the Royal Society Interface. The study of early warning indicators for biological transitions is an extremely active area of research. This paper reports the first study (to my knowledge) of the general trade-off between reliability and sensitivity of various indicators. Its introduction of an approximate method-based approach is a major contribution. The authors may wish to consider the following minor issues:

1. Line 41: “powerful way” is gratuitous

_"powerful" has been deleted_

2. Lines 61-63: I think some transcritical bifurcations can be “blue-sky” transitions; the claim should be qualified

_We apoloigize that the jargon is may not be used consistently in the field.  Transcritical bifurcations swap the stability of two fixed points, the fixed points to not appear/dissappear into the "blue".  We have removed the term_

3. Lines 74-76: I think the question of whether a noise-induced transition can be indicated is open. Particularly, the argument of Ditlevsen is based on a local approximation that is not expected to hold just prior to the transition; the author's may wish to qualify this claim. Also, “noise-induced” transition is ambiguous, here referring to the stochastic fluctuation between attractors, but more commonly referring to critical values in the magnitude of the fluctuations, as in Horsthemke and Lefever (1984).

_Excellent point, this has been clarified.  In particular, we note that the patterns put forward as warning signals for saddle node bifurcations, such as increasing variance or increasing autocorrelation, should not be expected to hold if the transition between attractors is driven purely by chance deviation._

4. Line 97: It is not obvious to be that autocorrelation from interpolation to a grid necessarily introduces artificial correlations. Either a reference or numerical example should be provided or the idea should be stated in a more conjectural mood.

5. Line 248: “This introduces a nuisance parameter known as the statistical significance criterion” is overly prejudicial. Differences of statistical philosophy are fine, but I think this goes over the line.

_This has been rephrased (we meant only in the technical sense as a parameter you must choose arbitrarily rather than estimate from data, but the point is well taken)_

6. Line 301: Typo

_The missing noun (citation to Dakos et al 2008) has been inserted,  thanks for catching this._

### 2012-04-21 boettiger.tex
corrections made based on reviews from Interface

However, for many ecological systems specific models are not available and general approaches are needed~\citep{Scheffer2009, Lade2012}
that do not depend on estimating the parameters of a known model of a specific system. 
This has lead to a variety of approaches based on summary statistics
~\citep[\emph{e.g.}][]{Carpenter2006, Held2004, Dakos2008, Guttal2008, Biggs2009, Carpenter2011, Seekell2011}
that look for generic signs of impending regime shifts.  
Here we extend earlier work by providing estimates 
of the ability of different potential indicators to accurately signal impending regime shifts,
and develop new approaches that both are more efficient
and also lay bare some of the important assumptions underlying attempts 
to find general warning signs of regime shifts.  
We distinguish this question from the extensive literature involving change-point analysis 
for the post-hoc identification of if and when a regime shift has occurred~\citep{Easterling1995, Rodionov2004, Lenton2009}, 
which is of little use if the goal is the advanced detection of the shift.

We begin by discussing the limitations of current approaches that rely on summary statistics
and provide a description of assumptions through the introduction of a model based approach to detect early warning signals.
We then illustrate how stochastic differential equation (SDE) models can be used
to reflect the uncertainty inherent in the detection of early warning signals. 
We caution against paradigms that are not useful for capturing uncertainty in
a model-selection based approach, such as information criteria.
Finally we use receiver-operating characteristics~\citep{Green1989, Keller2009}
as a powerful way to illustrate the sensitivity of different data sets
and different indicators have in detecting early warning signals and use this to explore a number of examples.
This approach provides a visualization of the types of errors that arise and how one can trade off between them.

\section{The summary statistics approach}
Foundational work on early warning signals has operated under the often-implicit assumption
that the system dynamics contain a saddle-node bifurcation by
looking for patterns that are associated with this kind of transition.
Typical patterns include an increasing trend in a summary statistic such as
variance~\citep{Carpenter2006}, autocorrelation~\citep{Held2004, Dakos2008},
skew~\citep{Guttal2008}, spectral ratio~\citep{Biggs2009}.
While attractive for their simplicity, such approaches must confront numerous challenges.
In this paper we argue for a model-based approach,
and describe how this can be done in a way that best addresses these difficulties.
We begin by enumerating several of the difficulties encountered in approaches lacking an explicit model.

\subsubsection*{Hidden assumptions}
The underlying assumption that the system contains a saddle-node bifurcation
can be easily overlooked in common summary-statistics based approaches.
For instance, variance may increase for reasons that
do not signal an approaching transition~\citep{Schreiber2003, Schreiber2008}.
Alternatively, variance may not increase as a bifurcation is approached~\citep{Livina2012, Dakos2011a}.
Some classes of sudden transitions may exhibit no warning signals~\cite{Hastings2010}.
Like saddle-node bifurcations, transcritical bifurcations involve an eigenvalue passing through zero,
and exhibit the patterns of critical slowing down and increased variance~\citep{Drake2010}.  
However, transcritical bifurcations are not ``blue-sky'' transitions that have made critical transitions so worrisome.% --
%they are characterized by a steady approach towards the alternative stable state, 
%such as a population size gradually approaching zero.  
While no approach will be applicable to all classes of sudden transitions,
it is certainly still useful to have an approach that detect transitions driven by
saddle node bifurcations, which have been found in many contexts~\citep[\emph{e.g.}, see][]{Scheffer2001}.

%Differing causes of the transition.
Even when we can exclude or ignore other dynamics and
restrict ourselves to systems that can produce a saddle-node bifurcation,
approaches based on critical slowing down or rising variance~\citep[\emph{e.g.}][]{Held2004, Scheffer2009, Carpenter2011}
must further assume that a changing parameter has brought the system closer to the bifurcation.
This assumption excludes at least three alternative explanations for the transition.
The first possibility is that a large perturbation of the system state
has moved the system into the alternative basin of attraction~\citep{Scheffer2001, Scheffer2001}.
This is an exogenous forcing that does not arise from the system dynamics, so it is not the kind of event we can expect to forecast.
(An example might be a sudden dramatic increase in fishing effort that pushes the population past a threshold.)
The second scenario is a purely noise-induced transition, a chance fluctuation that happens to carry the system across the boundary~\citep{Ditlevsen2010}.
As discussed in \citet{Livina2012}, such noise induced transitions cannot be predicted through early warning signals.
The third scenario is that the system does pass through a saddle-node bifurcation,
but rather than gradually and monotonically approaching the critical point, the
bifurcation parameter moves in a rapid or highly non-linear way, making the detection of any gradual trend impossible.

@@

\subsubsection*{Arbitrary windows}
In addition to the assumption of a saddle node bifurcation, 
the calculation of statistics that would be used to detect an impending transition is subject to several arbitrary choices.
A basic difficulty arises from the need to assume a time-series is ergodic: 
that averaging over time is equivalent to averaging over replicate realizations,
while trying to test if it is not.
Theoretically, the increasing trend in variance, autocorrelation, or other statistics
is something that would be measured across an ensemble -- across replicates. %% citation?
As true replicates are seldom available in systems for which developing warning signals would be most desirable,
these methods average across a single replicate using a moving window in time.
The selection of the size of this window and whether and by how much to overlap consecutive windows
varies across the literature.
\citet{Lenton2012} demonstrates that these differences can influence the results,
and that the different choices each carry advantages and disadvantages.

@@

\subsection*{Early warning signals as model choice}
It may be useful to think of the detection of early warning signals
as a problem of model choice rather than one of pattern recognition.
The model choice approach attempts to frame each of the possible scenarios as structurally different equations,
each with unknown parameters that must be estimated from the data.
In any model choice problem, it is important to identify the goal of the exercise --
such as the ability to generalize, to imitate reality, or to predict~\citep{Levins1966}.
In this case generality is more important than realism or predictive capability:
we will write down a general model that is capable of approximating
a wide class of models in which regime shifts are characterized by a saddle node bifurcation,
and a second generic model that is capable of representing the behavior of such systems
when they are not approaching a bifurcation.
These may be thought of as the hypothesis and null hypothesis,
though they are in fact compound hypotheses,
as we must first estimate the model parameters from the data.
In this approach it is not assumed that ``reality'' is included in the models being tested,
but that one of the models is a better approximation of the true dynamics than the other.
System whose dynamics violate the assumptions common to both models,
such as in the examples of ~\citet{Hastings2010} where systems exhibit sudden transitions without warning,
fall outside the set of cases where this approach would be valid;
though the inability of either model to match the system dynamics could be an indication of such a violation.

\subsection*{Models}
The normal form~\citep{Guckenheimer1983, Kuehn2011} for the saddle-node bifurcation is
\begin{equation}
\frac{\ud x}{\ud t} = r_t- x^2.
\label{saddle-node}
\end{equation}
where $x$ is the state variable and $r_t$ our bifurcation parameter,
which may be slowly varying in time and
consequently moving the system closer to a critical transition or regime shift~\citep{Scheffer2009}.
Transforming this canonical form to allow for an arbitrary mean in the state variable $\theta$,
the bifurcation looks like \( dx/dt = r_t- (\theta-x)^2 \), with fixed point \(\hat x = \sqrt{r_t} +\theta =: \phi(r_t)\).
We expand around the fixed point and express as a stochastic differential equation~\citep[\emph{e.g.}][]{Gardiner2009}:

@@

where $B_t$ is the standard Brownian motion. 
This expression captures the behavior of the system near the stable point as it approaches the bifurcation.
%It is also possible to derive this same expression from the master equation
%of various stochastic birth/death (or gains \& losses) models
%that contain saddle-node bifurcations, as we show in the appendix.
%Alternate dynamical forms are possible but will not differ substantially.

@@

\subsection*{Beyond hypothesis testing}
It is possible to frame the question of sensitivity, reliability, and adequate data in the language of hypothesis testing.
This introduces a nuisance parameter known as the statistical significance criterion.
In the hypothesis testing framework, a false positive is a Type I error,
which is defined relative to this arbitrary statistical significance criterion, most commonly 0.05.
By changing the criterion, one can increase or decrease the probability of the Type I error
at the cost of decreasing or increasing false negative or Type II error, which must also be defined relative to this criterion.

@@

Our second empirical data set comes from the glaciation record seen
in deuterium levels in Antarctic ice cores~\citep{Petit1999},
as analyzed by~\citet{Dakos2008}.
The data is preprocessed by linear interpolation and de-trending by Gaussian kernel smoothing
to be as consistent as possible with the original analysis.
We focus on the third glaciation event, consisting of 121 sample points.
The match is not exact since estimates the de-trending window size manually,
but the estimated correlations in the first-order auto-regression coefficients are in close agreement with that analysis.
De-trending is intended to make the data consistent with the assumptions of the warning signal detection~\citep{Dakos2008},
which did not apply to the other data sets~\citep{Drake2010}.
This time series is shown in the top panel of Figure~\ref{fig:glaciation}.
 \begin{figure}
   \begin{center}
     \includegraphics[width=.85\linewidth]{figures/GlaciationIII_warningsignal.pdf}
     \caption{A model-based calculation of warning signals for the Glaciation data analyzed in ~\citet{Dakos2008} (Glaciation III).  Panels as in Figure~\ref{fig:simulation}.}
     \label{fig:glaciation}
  \end{center}
 \end{figure}

\subsection*{Analysis}
The deviances $\delta$ observed are 5.1, 6.0, 83.9 for the simulation,
the chemostat data and the glaciation data, respectively.
Based on AIC score each is large enough to reject the null hypothesis of a stable model 
with it's one extra parameter, but this does not give the full picture of the anticipated error rates.  
The size of these differences reflects not only the magnitude of the difference in fit
between the models but also the arbitrary units of the raw likelihoods,
which are smaller for larger data-sets.
Consequently the glaciation score reflects as much the greater length of its time series as it does anything else.

>>>>

However, for many ecological systems specific models are not available and general approaches are needed~\citep{Scheffer2009, Lade2012}
that do not depend on estimating the parameters of a known model of a specific system. 
This has led to a variety of approaches based on summary statistics
~\citep[\emph{e.g.}][]{Carpenter2006, Held2004, Dakos2008, Guttal2008, Biggs2009, Carpenter2011, Seekell2011}
that look for generic signs of impending regime shifts.  
Here we extend earlier work by providing estimates 
of the ability of different potential indicators to accurately signal impending regime shifts,
and develop new approaches that both are more efficient
and also lay bare some of the important assumptions underlying attempts 
to find general warning signs of regime shifts.  
We distinguish this question from the extensive literature involving change-point analysis 
for the post-hoc identification of if and when a regime shift has occurred~\citep{Easterling1995, Rodionov2004, Lenton2009}, 
which is of little use if the goal is the advanced detection of the shift.

We begin by discussing the limitations of current approaches that rely on summary statistics
and provide a description of assumptions through the introduction of a model based approach to detect early warning signals.
We then illustrate how stochastic differential equation (SDE) models can be used
to reflect the uncertainty inherent in the detection of early warning signals. 
We caution against paradigms that are not useful for capturing uncertainty in
a model-selection based approach, such as information criteria.
Finally we use receiver-operating characteristics~\citep{Green1989, Keller2009}
as a way to illustrate the sensitivity of different data sets
and different indicators have in detecting early warning signals and use this to explore a number of examples.
This approach provides a visualization of the types of errors that arise and how one can trade off between them.

\section{The summary statistics approach}
Foundational work on early warning signals has operated under the often-implicit assumption
that the system dynamics contain a saddle-node bifurcation by
looking for patterns that are associated with this kind of transition.
A saddle-node bifurcation is the result of a stable node and an unstable
saddle-point colliding and annihilating, leaving the system to transition 
to another attractor~\citep[see ][or other dynamical systems text 
for an introduction to this and other bifurcations]{Guckenheimer1983}.

Typical patterns include an increasing trend in a summary statistic such as
variance~\citep{Carpenter2006}, autocorrelation~\citep{Held2004, Dakos2008},
skew~\citep{Guttal2008}, spectral ratio~\citep{Biggs2009}.
While attractive for their simplicity, such approaches must confront numerous challenges.
In this paper we argue for a model-based approach,
and describe how this can be done in a way that best addresses these difficulties.
We begin by enumerating several of the difficulties encountered in approaches lacking an explicit model.

\subsubsection*{Hidden assumptions}
The underlying assumption that the system contains a saddle-node bifurcation
can be easily overlooked in common summary-statistics based approaches.
For instance, variance may increase for reasons that
do not signal an approaching transition~\citep{Schreiber2003, Schreiber2008}.
Alternatively, variance may not increase as a bifurcation is approached~\citep{Livina2012, Dakos2011a}.
Some classes of sudden transitions may exhibit no warning signals~\cite{Hastings2010}.
Like saddle-node bifurcations, transcritical bifurcations involve an eigenvalue passing through zero,
and exhibit the patterns of critical slowing down and increased variance~\citep{Drake2010}.  
However, transcritical bifurcations involve a change in stability of a fixed point, rather
than the sudden dissapearance of a fixed point that has made critical transitions so worrisome.% --
%they are characterized by a steady approach towards the alternative stable state, 
%such as a population size gradually approaching zero.  
While no approach will be applicable to all classes of sudden transitions,
it is certainly still useful to have an approach that detect transitions driven by
saddle-node bifurcations, which have been found in many contexts~\citep[\emph{e.g.}, see][]{Scheffer2001}.

%Differing causes of the transition.
Even when we can exclude or ignore other dynamics and
restrict ourselves to systems that can produce a saddle-node bifurcation,
approaches based on critical slowing down or rising variance~\citep[\emph{e.g.}][]{Held2004, Scheffer2009, Carpenter2011}
must further assume that a changing parameter has brought the system closer to the bifurcation.
This assumption excludes at least three alternative explanations for the transition.
The first possibility is that a large perturbation of the system state
has moved the system into the alternative basin of attraction~\citep{Scheffer2001, Scheffer2001}.
This is an exogenous forcing that does not arise from the system dynamics, so it is not the kind of event we can expect to forecast.
(An example might be a sudden dramatic increase in fishing effort that pushes the population past a threshold.)
The second scenario is a purely noise-induced transition, a chance fluctuation that happens to carry the system across the boundary~\citep{Ditlevsen2010}.
\citet{Livina2012} indicate that such noise induced transitions cannot be predicted through early warning signals -- 
at least they are not expected to exhibit the same early warning patterns of increased variance and increased autocorrelation
anticipated in the case of a saddle-node bifurcation.
The third scenario is that the system does pass through a saddle-node bifurcation,
but rather than gradually and monotonically approaching the critical point, the
bifurcation parameter moves in a rapid or highly non-linear way, making the detection of any gradual trend impossible.

@@

\subsubsection*{Arbitrary windows}
In addition to the assumption of a saddle-node bifurcation, 
the calculation of statistics that would be used to detect an impending transition is subject to several arbitrary choices.
A basic difficulty arises from the need to assume a time-series is ergodic: 
that averaging over time is equivalent to averaging over replicate realizations,
while trying to test if it is not.
Theoretically, the increasing trend in variance, autocorrelation, or other statistics
is something that would be measured across an ensemble -- across replicates. %% citation?
As true replicates are seldom available in systems for which developing warning signals would be most desirable,
these methods average across a single replicate using a moving window in time.
The selection of the size of this window and whether and by how much to overlap consecutive windows
varies across the literature.
\citet{Lenton2012} demonstrates that these differences can influence the results,
and that the different choices each carry advantages and disadvantages.

@@

\subsection*{Early warning signals as model choice}
It may be useful to think of the detection of early warning signals
as a problem of model choice rather than one of pattern recognition.
The model choice approach attempts to frame each of the possible scenarios as structurally different equations,
each with unknown parameters that must be estimated from the data.
In any model choice problem, it is important to identify the goal of the exercise --
such as the ability to generalize, to imitate reality, or to predict~\citep{Levins1966}.
In this case generality is more important than realism or predictive capability:
we will write down a general model that is capable of approximating
a wide class of models in which regime shifts are characterized by a saddle-node bifurcation,
and a second generic model that is capable of representing the behavior of such systems
when they are not approaching a bifurcation.
These may be thought of as the hypothesis and null hypothesis,
though they are in fact compound hypotheses,
as we must first estimate the model parameters from the data.
In this approach it is not assumed that ``reality'' is included in the models being tested,
but that one of the models is a better approximation of the true dynamics than the other.
System whose dynamics violate the assumptions common to both models,
such as in the examples of ~\citet{Hastings2010} where systems exhibit sudden transitions without warning,
fall outside the set of cases where this approach would be valid;
though the inability of either model to match the system dynamics could be an indication of such a violation.

\subsection*{Models}
The normal form~\citep{Guckenheimer1983, Kuehn2011} for the saddle-node bifurcation is
\begin{equation}
\frac{\ud x}{\ud t} = r_t- x^2.
\label{saddle-node}
\end{equation}
where $x$ is the state variable and $r_t$ our bifurcation parameter.
We have added a subscript $t$ to the bifurcation parameter as a reminder that
it is the value which may be slowly varying in time and
consequently moving the system closer to a critical transition or regime shift~\citep{Scheffer2009}.
Transforming this canonical form to allow for an arbitrary mean in the state variable $\theta$,
the bifurcation looks like \( dx/dt = r_t- (\theta-x)^2 \), with fixed point \(\hat x = \sqrt{r_t} +\theta =: \phi(r_t)\).
We expand around the fixed point and express as a stochastic differential equation~\citep[\emph{e.g.}][]{Gardiner2009}:

@@

where $B_t$ is the standard Brownian motion. 
This expression captures the behavior of the system near the stable point as it approaches the bifurcation.
Allowing the stochastic term to scale with the square root of $\phi$ follows from the assumption that
of an internal-noise process, such as demographic stochasticity, that arises in deriving the SDE from
a Markov process, see~\citet{Kampen2007a} or~\citet{Black2012}.  
%It is also possible to derive this same expression from the master equation
%of various stochastic birth/death (or gains \& losses) models
%that contain saddle-node bifurcations, as we show in the appendix.
%Alternate dynamical forms are possible but will not differ substantially.

@@

\subsection*{Beyond hypothesis testing}
It is possible to frame the question of sensitivity, reliability, and adequate data in the language of hypothesis testing.
This introduces the need for selecting a statistical significance criterion.
In the hypothesis testing framework, a false positive is a Type I error,
which is defined relative to this arbitrary statistical significance criterion, most commonly 0.05.
By changing the criterion, one can increase or decrease the probability of the Type I error
at the cost of decreasing or increasing false negative or Type II error, which must also be defined relative to this criterion.

@@

Our second empirical data set comes from the glaciation record seen
in deuterium levels in Antarctic ice cores~\citep{Petit1999},
as analyzed by~\citet{Dakos2008}.
The data are preprocessed by linear interpolation and de-trending by Gaussian kernel smoothing
to be as consistent as possible with the original analysis.
We focus on the third glaciation event, consisting of 121 sample points.
The match is not exact since~\citet{Dakos2008} estimates the de-trending window size manually,
but the estimated correlations in the first-order auto-regression coefficients are in close agreement with that analysis.
De-trending is intended to make the data consistent with the assumptions of the warning signal detection~\citep{Dakos2008},
which did not apply to the other data sets~\citep{Drake2010}.
This time series is shown in the top panel of Figure~\ref{fig:glaciation}.
 \begin{figure}
   \begin{center}
     \includegraphics[width=.85\linewidth]{figures/GlaciationIII_warningsignal.pdf}
     \caption{A model-based calculation of warning signals for the Glaciation data analyzed in ~\citet{Dakos2008} (Glaciation III).  Panels as in Figure~\ref{fig:simulation}.}
     \label{fig:glaciation}
  \end{center}
 \end{figure}

\subsection*{Analysis}
The deviances $\delta$ observed are 5.1, 6.0, 83.9 for the simulation,
the chemostat data and the glaciation data, respectively.
Based on AIC score each is large enough to reject the null hypothesis of a stable model 
with its one extra parameter, but this does not give the full picture of the anticipated error rates.  
The size of these differences reflects not only the magnitude of the difference in fit
between the models but also the arbitrary units of the raw likelihoods,
which are smaller for larger data-sets.
Consequently the glaciation score reflects as much the greater length of its time series as it does anything else.
